---
title: "Story Pitch — Yasmeen Saadi, 913-208-6316"
output: html_document
date: "2025-06-02"
---
#rsw comments below -
Yasmeen - This is a very thoughtful proposal with a lot of detail. You have focused it well and spelled out a good plan. The next steps will be to build on this by adding the code to illustrate and expand on the points here. See my examples below. Today we're focused on data visualizations, so you'll be able to add those here.

In the late 1900s, Black enrollment at the City Colleges of Chicago was substantially higher than any other group of race/ethnicity, according to data from the Illinois Community College Board. By the 2000s, enrollment numbers between Black and Hispanic students started to even out.

However, in the past 20 years, enrollment for Black students has steadily decreased as Hispanic enrollment has stayed relatively steady.

Using data from the City Colleges of Chicago (CCC) system, I looked into how community college enrollment by race/ethnicity has changed over the past 20 years in the city. This data, along with interviews and historical context, could help us understand how the accessibility/use of these colleges for students of color in the area has developed.

Especially as DEI initiatives and grants for students of color are being threatened, it is timely to take a closer look at the institutions that serve as a more affordable option and if they are truly serving the students who need them.

Community colleges are especially important for low-income students and those seeking a more affordable way to complete their degrees. In each of CCC's seven colleges, students of color make up over 60% of the student enrollment. By compiling the data from the past 10 years, we can see if the percentages of Black and Latinx students have always been high or if they have increased over the years. The Investigative Project on Race and Equity was originally interested in looking at acceptances and rejections across zip codes. However, all seven colleges are open enrollment, meaning anyone who applies is awarded admission.

Just by looking at the statistical digests from FY 2017 and FY 2023, over the six years, the enrollment numbers for Black students has decreased slightly, while the numbers for Latinx students have increased. In the 2022-23 academic year, which is the most recent year available, 7,722 students enrolled in a degree program. Of those, 27.5% were Black and 48.1% were Latinx. In 2017, the degree enrollment was similar at 7,224 students. However, 33.4% were Black and 37.7% were Latinx. I could continue making these comparisons to see if this has been a steady trend.
#rsw comment - Great - let's make a chart from this data. Today you will be learning data visualizations - perfect!

CCC was founded in 1911. For a more personal angle, we could try to find older POC alumni who attended the colleges years ago to compare to the experiences of students of color now. Students who have attended CCC would care about the article, especially if they have feelings on the cultural/racial representation within the colleges. We could do this by searching for relevant Facebook groups or directories.
#rsw comment - great idea to contact the alumni networks

##Records/data 

City Colleges of Chicago includes a statistical digest in its budget each fiscal year which includes numbers on enrollment by degree area and ethnicity. The only issue is that there are many other students who take courses through the community colleges who are not on a degree-seeking course, who I do not believe are represented in this data.
#rsw comment - good point. back in the day, i took a few community college classes but they were not for a degree. contact the colleges to see how you can fill this gap, if at all

I plan to go through the budget documents to see if there is more comprehensive enrollment data for non-degree-seeking students.

College Scorecard also provides snapshots on each community college, including graduation rate, annual costs and median household income, which can provide good context.

#rsw comment - the college scorecards can be simple profile graphics for each of the schools. it would be worth checking if income changes over time

##People I can interview 

- Lorenzo Baber is the director of UIUC's Office of Community College Research and Leadership. He or someone on the research team could provide insight on persisting equity issues within community colleges.

- In April, Chicago Public Schools, CCC and the University of Illinois Chicago launched “Chicago Roadmap 2.0,” which will try to create a pipeline for students seeking higher education. The website does not include the directors of the project, but I could reach out to the general email to have someone on the project speak about the local need for students in the city.

- Young and old alumni of the colleges who are people of color could show the range of student experiences.

##Visuals

#rsw comment: great ideas

One graphic idea is a demographic map of Chicago with pinpoints for each of the seven community colleges.

Another graphic idea is a bar graph where you could scroll throughout the years and look at the enrollment by ethnicity for Black and Latinx students for the last 10 years.

If it would be possible to find real people who had gone through a City of Columbia College years ago, it would be cool to have photos of them in college and them now.
#rsw comment - yearbooks, if available, would be a good source

##Realistic deadline 

I would likely need two weeks to finish this project in order to analyze my code and have enough time to schedule/conduct interviews.


#rsw comment - add your data and discuss it as part of the narrative


# Load software libraries
```{r}
#install.packages("tidyverse")
#install.packages("janitor")
#install.packages("readxl")
#install.packages("rvest")
```

### Load libraries
```{r}
library(tidyverse)
library(janitor)
library(readxl)
library(rvest)

```


# Load your data

# Explore the data types
```{r}

#glimpse(YOUR_DATA)

#glimpse(data)


```

Data details:

# Percent of whole by race

```{r}
# 
#data <- data |>
#  mutate(asian = (as.numeric(asian)),
#    asian_pct = (asian/total)*100,
#        black_pct = (black/total)*100,
#         latinx_pct = (latinx/total)*100,
#         white_pct = (white/total)*100,
#         other_pct = (other/total)*100)
```
#Fall enrollment data by race since 2000
```{r}
enroll <- read.csv("./data/Fall Enrollment CCC Since 2000 - Sheet1 (1).csv") |>
  janitor::clean_names()
```
#Q: What is the percent change in fall enrollment between 2024 and 2014?
#A: -25.2% for all City Colleges of Chicago
```{r}
library(tidyverse)
enroll1 <- enroll |>
  select(fall_year, college, total) |>
  filter(fall_year == "2024" | fall_year == "2014") |>
  pivot_wider(names_from = fall_year, values_from = total, id_cols = college) |>
  janitor::clean_names() |>
 mutate(x2024 = as.numeric(str_replace_all(x2024, ",", ""))) |> 
  mutate(x2014 = as.numeric(str_replace_all(x2014, ",", ""))) |> 
  mutate(pct_change = (x2024-x2014)/x2014)

  
```
#Q: What is the percent change in fall enrollment between 2024 and 2004 for Black students?
#A: -50.2% for all City Colleges of Chicago
```{r}
library(tidyverse)
black_enrollment_20years <- enroll |>
  select(fall_year, college, african_american_non_latino) |>
  filter(fall_year == "2024" | fall_year == "2004") |>
  pivot_wider(names_from = fall_year, values_from = african_american_non_latino, id_cols = college) |>
  janitor::clean_names() |>
 mutate(x2024 = as.numeric(str_replace_all(x2024, ",", ""))) |> 
  mutate(x2004 = as.numeric(str_replace_all(x2004, ",", ""))) |> 
  mutate(pct_change = (x2024-x2004)/x2004)

```
#Q: What is the percent change in fall enrollment between 2024 and 2014 for Black students?
#A: 
```{r}
library(tidyverse)
black_enrollment_10years <- enroll |>
  select(fall_year, college, african_american_non_latino) |>
  filter(fall_year == "2024" | fall_year == "2014") |>
  pivot_wider(names_from = fall_year, values_from = african_american_non_latino, id_cols = college) |>
  janitor::clean_names() |>
 mutate(x2024 = as.numeric(str_replace_all(x2024, ",", ""))) |> 
  mutate(x2014 = as.numeric(str_replace_all(x2014, ",", ""))) |> 
  mutate(pct_change = (x2024-x2014)/x2014)

```
#Q: What is the percent change in fall enrollment between 2024 and 2014 for Latinx students?
#A: -3.3% change
```{r}
library(tidyverse)
latinx_enrollment_10years <- enroll |>
  select(fall_year, college, latino) |>
  filter(fall_year == "2024" | fall_year == "2014") |>
  pivot_wider(names_from = fall_year, values_from = latino, id_cols = college) |>
  janitor::clean_names() |>
 mutate(x2024 = as.numeric(str_replace_all(x2024, ",", ""))) |> 
  mutate(x2014 = as.numeric(str_replace_all(x2014, ",", ""))) |> 
  mutate(pct_change = (x2024-x2014)/x2014)

```
#Q: What is the percent change in fall enrollment between 2024 and 2004 for Latinx students?
#A: -7.3%
```{r}
library(tidyverse)
latinx_enrollment_20years <- enroll |>
  select(fall_year, college, latino) |>
  filter(fall_year == "2024" | fall_year == "2004") |>
  pivot_wider(names_from = fall_year, values_from = latino, id_cols = college) |>
  janitor::clean_names() |>
 mutate(x2024 = as.numeric(str_replace_all(x2024, ",", ""))) |> 
  mutate(x2004 = as.numeric(str_replace_all(x2004, ",", ""))) |> 
  mutate(pct_change = (x2024-x2004)/x2004)

```
#Q: How has the makeup of 
#A: -7.3%

<iframe title="Black student enrollment in City Colleges of Chicago drops as Latinx student enrollment stays steady " aria-label="Interactive line chart" id="datawrapper-chart-iHTUw" src="https://datawrapper.dwcdn.net/iHTUw/1/" scrolling="no" frameborder="0" style="width: 0; min-width: 100% !important; border: none;" height="400" data-external="1"></iframe><script type="text/javascript">!function(){"use strict";window.addEventListener("message",(function(a){if(void 0!==a.data["datawrapper-height"]){var e=document.querySelectorAll("iframe");for(var t in a.data["datawrapper-height"])for(var r,i=0;r=e[i];i++)if(r.contentWindow===a.source){var d=a.data["datawrapper-height"][t]+"px";r.style.height=d}}}))}();
</script>

#rsw comment - add percent of whokle graphic here... etc.

