---
title: "How City Colleges of Chicago fall enrollment has changed since 2000 — Yasmeen Saadi"
output: html_document
date: "2025-06-04"
---

<iframe title="City Colleges of Chicago" aria-label="Locator map" id="datawrapper-chart-3LN2l" src="https://datawrapper.dwcdn.net/3LN2l/2/" scrolling="no" frameborder="0" style="width: 0; min-width: 100% !important; border: none;" height="563" data-external="1"></iframe><script type="text/javascript">!function(){"use strict";window.addEventListener("message",(function(a){if(void 0!==a.data["datawrapper-height"]){var e=document.querySelectorAll("iframe");for(var t in a.data["datawrapper-height"])for(var r,i=0;r=e[i];i++)if(r.contentWindow===a.source){var d=a.data["datawrapper-height"][t]+"px";r.style.height=d}}}))}();
</script>

Excluding a recent rise in overall enrollment after the COVID-19 pandemic — in the past 15 years, enrollment for Black students has steadily decreased as Hispanic enrollment has stayed relatively steady. This has resulted in a student body with a heavier Hispanic/Latinx presence.

In the 1980s, Black enrollment at the City Colleges of Chicago was substantially higher than any other group of race/ethnicity, according to data from the Illinois Community College Board. By the late 1990s/early 2000s, enrollment numbers between Black and Hispanic students started to even out. 2012 was the crossing point. After 2012, Black enrollment declined lower than than the Hispanic enrollment.

Combining this data with interviews and historical context can help us understand how the accessibility/use of these colleges for students of color in the area has developed.

<iframe title="Fall enrollment across City Colleges of Chicago for Black and Hispanic students" aria-label="Grouped Columns" id="datawrapper-chart-xxniH" src="https://datawrapper.dwcdn.net/xxniH/1/" scrolling="no" frameborder="0" style="width: 0; min-width: 100% !important; border: none;" height="517" data-external="1"></iframe><script type="text/javascript">!function(){"use strict";window.addEventListener("message",(function(a){if(void 0!==a.data["datawrapper-height"]){var e=document.querySelectorAll("iframe");for(var t in a.data["datawrapper-height"])for(var r,i=0;r=e[i];i++)if(r.contentWindow===a.source){var d=a.data["datawrapper-height"][t]+"px";r.style.height=d}}}))}();
</script>

<iframe title="Changes in Black student enrollment compared to college totals" aria-label="Grouped Columns" id="datawrapper-chart-Zp5aq" src="https://datawrapper.dwcdn.net/Zp5aq/2/" scrolling="no" frameborder="0" style="width: 0; min-width: 100% !important; border: none;" height="473" data-external="1"></iframe><script type="text/javascript">!function(){"use strict";window.addEventListener("message",(function(a){if(void 0!==a.data["datawrapper-height"]){var e=document.querySelectorAll("iframe");for(var t in a.data["datawrapper-height"])for(var r,i=0;r=e[i];i++)if(r.contentWindow===a.source){var d=a.data["datawrapper-height"][t]+"px";r.style.height=d}}}))}();
</script>

ADD ACADEMIC CONTEXT ABOUT THE POSITION OF COMMUNITY COLLEGES — STUDY

Study by ____ looking at community colleges as racialized institutions. most accessible postsecondary pathway for minoritized students is also the most chronically under-resourced sector of higher education in the United States.

Especially as DEI initiatives and grants for students of color are being threatened, it is timely to take a closer look at the institutions that serve as a more affordable option and if they are truly serving the students who need them. We can also compare CCC's growth with the population growth of the city to see if the two align. If not, we would need to dig in to learn how come.

Community colleges are especially important for low-income students and those seeking a more affordable way to complete their degrees. In each of CCC's seven colleges, students of color make up over 60% of the student enrollment. By compiling the data from the past 10 years, we can see if the percentages of Black and Latinx students have always been high or if they have increased over the years. The Investigative Project on Race and Equity was originally interested in looking at acceptances and rejections across zip codes. However, all seven colleges are open enrollment, meaning anyone who applies is awarded admission.

<iframe title="Black student enrollment in City Colleges of Chicago drops as Latinx student enrollment stays steady" aria-label="Interactive line chart" id="datawrapper-chart-iHTUw" src="https://datawrapper.dwcdn.net/iHTUw/2/" scrolling="no" frameborder="0" style="width: 0; min-width: 100% !important; border: none;" height="400" data-external="1"></iframe><script type="text/javascript">!function(){"use strict";window.addEventListener("message",(function(a){if(void 0!==a.data["datawrapper-height"]){var e=document.querySelectorAll("iframe");for(var t in a.data["datawrapper-height"])for(var r,i=0;r=e[i];i++)if(r.contentWindow===a.source){var d=a.data["datawrapper-height"][t]+"px";r.style.height=d}}}))}();
</script>

Just by looking at the statistical digests from FY 2017 and FY 2023, over the six years, the enrollment numbers for Black students has decreased slightly, while the numbers for Latinx students have increased. In the 2022-23 academic year, which is the most recent year available, 7,722 students enrolled in a degree program. Of those, 27.5% were Black and 48.1% were Latinx. In 2017, the degree enrollment was similar at 7,224 students. However, 33.4% were Black and 37.7% were Latinx. I could continue making these comparisons to see if this has been a steady trend.

Because CCC has a long history, for a more personal angle, we could try to find older POC alumni who attended the colleges years ago to compare to the experiences of students of color now. Students who have attended CCC would care about the article, especially if they have feelings on the cultural/racial representation within the colleges. We could do this by searching for relevant Facebook groups or directories.

## Records/data 

The Illinois Community College Board has data on fall enrollment available dating back to 1983. I compiled enrollment data by ethnicity from 2000 to 2024 to better understand how enrollment has changed.

This data was much more comprehensive than the data provided in City Colleges of Chicago's statistical digests. It included students who were not seeking degrees, which the statistical digests did not. However, my findings on the decline in Black enrollment and steadiness of Latinx enrollment were the same for both.

College Scorecard also provides snapshots on each community college, including graduation rate, annual costs and median household income, which is good context. I used this in my first graphic to provide general overviews of each college.

## People I can interview 

- Lorenzo Baber is the director of UIUC's Office of Community College Research and Leadership. He or someone on the research team could provide insight on persisting equity issues within community colleges.

Report on racially equitable practices on community colleges: nearly half of all students of color enrolled in U.S. postsecondary education attend a community college (American Association of Community Colleges, 2022; Long, 2016), decades of research suggest that the egalitarian principles of community colleges do not necessarily translate to racially equitable student experiences or outcomes 

minoritized students disproportionately attend community colleges, therefore they are critical sites for intervening in racial inequity

- In April, Chicago Public Schools, CCC and the University of Illinois Chicago launched “Chicago Roadmap 2.0,” which will try to create a pipeline for students seeking higher education. The website does not include the directors of the project, but I could reach out to the general email to have someone on the project speak about the local need for students in the city.

- A researcher/historian at the Abakanowicz Research Center within the Chicago History Museum could provide context on the demographic development of the city and the rise in Latinx population.

- Young and old alumni of the colleges who are people of color could show the range of student experiences.

## Other visuals

I would like to insert a chloropleth map of the Hispanic population of Chicago in 2020, 2010 and 2000. I am just not sure how to get the census data to generate on the map. I could also do a chloropleth map showing changes in household income over time.

If it would be possible to find real people who had gone through a City of Columbia College years ago, it would be cool to have photos of them in college and them now. I could look through yearbooks to find sources.

## Realistic deadline 

I would likely need two weeks to finish this project in order to analyze my code and have enough time to schedule/conduct interviews.

## Out migration of Black families? The rise of other institutions?

Looking at CPS enrollment — https://kidsfirstchicago.org/publications/enrollment-crisis

Correlation to Chicago Public Schools — Not only have Black students recorded the largest year-over-year decrease in enrollment but also the largest decrease since 2014. There are 44,680 fewer Black students enrolled in CPS this year than 10 years prior. That marks a decline of 29% during 10 years. — https://www.illinoispolicy.org/chicago-public-schools-sees-uptick-in-enrollment-but-still-lower-than-10-years-ago/


Increases in Hispanic enrollment and population growth — impacting community colleges and K-12 public schools — do we have the English learning resources to support this growth? — https://wgntv.com/migrants-in-chicago/chicago-public-school-teachers-allegedly-told-to-pass-migrant-students-regardless-of-grades-report/




### Load software libraries
```{r}
#install.packages("tidyverse")
#install.packages("janitor")
#install.packages("readxl")
#install.packages("rvest")
```

```{r}
library(tidyverse)
library(janitor)
library(readxl)
library(rvest)

```


### Loading data:

```{r}
enroll <- read.csv("../data/Fall Enrollment CCC Since 2000 - Sheet1 (1).csv") |>
  janitor::clean_names()
```

### Data analysis:

Let's first take a look at the breakdown of students by college:
```{r}
library(tidyverse)
library(janitor)
library(knitr)

# Fall 2024 breakdown of enrollment by college and ethnicity (wide format)
fall_2024_breakdown <- enroll |>
  filter(fall_year == "2024") |>
  select(
    college,
    asian,
    native_american,
    african_american_non_latino,
    latino,
    white_non_latino,
    u_s_nonresident,
    pacific_islander,
    middle_eastern_or_north_african,
    two_or_more_races,
    all_other_or_no_indication,
    total
  ) |>
  mutate(across(
    .cols = -college,
    .fns = ~ as.numeric(str_replace_all(as.character(.), ",", ""))
  )) |>
  clean_names()

# Show the table
kable(fall_2024_breakdown, caption = "Fall 2024 Enrollment by College and Ethnicity")

```
This table shows that for six of the colleges, African American and Latino students have the highest enrollment. The only outlier is Wilbur Wright.

Using this information, we can look at the percentages of Black and Latino students. It would also be interesting to know if this breakdown has always been the case. 
```{r}
enroll <- enroll |>
  mutate(
    asian = as.numeric(str_replace_all(asian, ",", "")),
    native_american = as.numeric(str_replace_all(native_american, ",", "")),
    african_american_non_latino = as.numeric(str_replace_all(african_american_non_latino, ",", "")),
    latino = as.numeric(str_replace_all(latino, ",", "")),
    white_non_latino = as.numeric(str_replace_all(white_non_latino, ",", "")),
    total = as.numeric(str_replace_all(total, ",", "")),
    asian_pct = (asian / total) * 100,
    native_american_pct = (native_american / total) * 100,
    african_american_non_latino_pct = (african_american_non_latino / total) * 100,
    latino_pct = (latino / total) * 100,
    white_non_latino_pct = (white_non_latino / total) * 100
  )

```
Let's look at how the percentage of Black students in each college has changed since 2014:
```{r}
library(tidyverse)
makeup_black <- enroll |>
  select(fall_year, college, african_american_non_latino_pct) |>
  filter(fall_year == "2024" | fall_year == "2014") |>
  pivot_wider(names_from = fall_year, values_from = african_american_non_latino_pct, id_cols = college) |>
  janitor::clean_names() |>
 mutate(x2024 = as.numeric(str_replace_all(x2024, ",", ""))) |> 
  mutate(x2014 = as.numeric(str_replace_all(x2014, ",", ""))) |> 
  mutate(difference = (x2024-x2014))

# Show the table
kable(makeup_black, caption = "Breakdown of Black students in 2014 and 2024 by percentage")
```

We can take it even further back to look at how the percentage of Black students in each college has changed since 2004:
```{r}
library(tidyverse)
makeup_black_2004 <- enroll |>
  select(fall_year, college, african_american_non_latino_pct) |>
  filter(fall_year == "2024" | fall_year == "2004") |>
  pivot_wider(names_from = fall_year, values_from = african_american_non_latino_pct, id_cols = college) |>
  janitor::clean_names() |>
 mutate(x2024 = as.numeric(str_replace_all(x2024, ",", ""))) |> 
  mutate(x2014 = as.numeric(str_replace_all(x2004, ",", ""))) |> 
  mutate(difference = (x2024-x2004))

# Show the table
kable(makeup_black_2004, caption = "Breakdown of Black students in 2004 and 2024 by percentage")
```
^I am not sure why the above table also shows a 2014 column with the a copy of the 2004 data.

Now let's look at the change in the percentage of Latinx student since 2004 across each college:
```{r}
library(tidyverse)
makeup_latinx_2004 <- enroll |>
  select(fall_year, college, latino_pct) |>
  filter(fall_year == "2024" | fall_year == "2004") |>
  pivot_wider(names_from = fall_year, values_from = latino_pct, id_cols = college) |>
  janitor::clean_names() |>
 mutate(x2024 = as.numeric(str_replace_all(x2024, ",", ""))) |> 
  mutate(x2004 = as.numeric(str_replace_all(x2004, ",", ""))) |> 
  mutate(difference = (x2024-x2004))

# Show the table
kable(makeup_latinx_2004, caption = "Breakdown of Latinx students in 2004 and 2024 by percentage")
```

The results show that excluding Harry S Truman College, the percentage of Latinx students who make up the student body have increased an average of 19.6% across the other six between 2004 and 2024.
##### Q: What was the percent change in fall enrollment between 2024 and 2014 for CCC as a whole?

I asked this question because I wanted to know if overall enrollment dropping could be a reason for the large drops in Black student enrollment.

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(knitr)

# Black student enrollment change
black_enrollment_10years <- enroll |>
  select(fall_year, college, african_american_non_latino) |>
  filter(fall_year == "2024" | fall_year == "2014") |>
  pivot_wider(names_from = fall_year, values_from = african_american_non_latino, id_cols = college) |>
  clean_names() |>
  mutate(
    x2024 = as.numeric(str_replace_all(x2024, ",", "")),
    x2014 = as.numeric(str_replace_all(x2014, ",", "")),
    pct_change = (x2024 - x2014) / x2014
  )

# Show the black enrollment table
kable(black_enrollment_10years, caption = "Change in Black Enrollment by College")

# Overall enrollment change
enroll1 <- enroll |>
  select(fall_year, college, total) |>
  filter(fall_year == "2024" | fall_year == "2014") |>
  pivot_wider(names_from = fall_year, values_from = total, id_cols = college) |>
  clean_names() |>
  mutate(
    x2024 = as.numeric(str_replace_all(x2024, ",", "")),
    x2014 = as.numeric(str_replace_all(x2014, ",", "")),
    pct_change = (x2024 - x2014) / x2014
  )

# Show the total enrollment table
kable(enroll1, caption = "Change in Total Enrollment by College")
```

```

#Q: What is the percent change in fall enrollment between 2024 and 2004 for Black students?
#A: 50.2% decrease for all City Colleges of Chicago

```{r}
library(tidyverse)
black_enrollment_20years <- enroll |>
  select(fall_year, college, african_american_non_latino) |>
  filter(fall_year == "2024" | fall_year == "2004") |>
  pivot_wider(names_from = fall_year, values_from = african_american_non_latino, id_cols = college) |>
  janitor::clean_names() |>
 mutate(x2024 = as.numeric(str_replace_all(x2024, ",", ""))) |> 
  mutate(x2004 = as.numeric(str_replace_all(x2004, ",", ""))) |> 
  mutate(pct_change = (x2024-x2004)/x2004)
  
  # Show the total enrollment table
kable(black_enrollment_20years, caption = "Percent change in fall enrollment between 2024 and 2004 for Black students")

```
#Q: What is the percent change in fall enrollment between 2024 and 2014 for Black students?
```{r}
library(tidyverse)
black_enrollment_10years <- enroll |>
  select(fall_year, college, african_american_non_latino) |>
  filter(fall_year == "2024" | fall_year == "2014") |>
  pivot_wider(names_from = fall_year, values_from = african_american_non_latino, id_cols = college) |>
  janitor::clean_names() |>
 mutate(x2024 = as.numeric(str_replace_all(x2024, ",", ""))) |> 
  mutate(x2014 = as.numeric(str_replace_all(x2014, ",", ""))) |> 
  mutate(pct_change = (x2024-x2014)/x2014)
  
  # Show the total enrollment table
kable(black_enrollment_10years, caption = "Percent change in fall enrollment between 2024 and 2014 for Black students")

```
#Q: What is the percent change in fall enrollment between 2024 and 2014 for Latinx students?
#A: negative 3.3% change
```{r}
library(tidyverse)
latinx_enrollment_10years <- enroll |>
  select(fall_year, college, latino) |>
  filter(fall_year == "2024" | fall_year == "2014") |>
  pivot_wider(names_from = fall_year, values_from = latino, id_cols = college) |>
  janitor::clean_names() |>
 mutate(x2024 = as.numeric(str_replace_all(x2024, ",", ""))) |> 
  mutate(x2014 = as.numeric(str_replace_all(x2014, ",", ""))) |> 
  mutate(pct_change = (x2024-x2014)/x2014)
  
  # Show the total enrollment table
kable(latinx_enrollment_10years, caption = "Percent change in fall enrollment between 2024 and 2014 for Latinx students")

```
#Q: What is the percent change in fall enrollment between 2024 and 2004 for Latinx students?
#A: negative 7.3% change
```{r}
library(tidyverse)
latinx_enrollment_20years <- enroll |>
  select(fall_year, college, latino) |>
  filter(fall_year == "2024" | fall_year == "2004") |>
  pivot_wider(names_from = fall_year, values_from = latino, id_cols = college) |>
  janitor::clean_names() |>
 mutate(x2024 = as.numeric(str_replace_all(x2024, ",", ""))) |> 
  mutate(x2004 = as.numeric(str_replace_all(x2004, ",", ""))) |> 
  mutate(pct_change = (x2024-x2004)/x2004)
  
    # Show the total enrollment table
kable(latinx_enrollment_20years, caption = "Percent change in fall enrollment between 2024 and 2004 for Latinx students")

```


#rsw comment - add percent of whole graphic here... etc.

