---
title: "How City Colleges of Chicago fall enrollment has changed since 2000"
output: html_document
date: "2025-06-04 — Yasmeen Saadi"
---

<iframe title="City Colleges of Chicago" aria-label="Locator map" id="datawrapper-chart-3LN2l" src="https://datawrapper.dwcdn.net/3LN2l/2/" scrolling="no" frameborder="0" style="width: 0; min-width: 100% !important; border: none;" height="563" data-external="1"></iframe><script type="text/javascript">!function(){"use strict";window.addEventListener("message",(function(a){if(void 0!==a.data["datawrapper-height"]){var e=document.querySelectorAll("iframe");for(var t in a.data["datawrapper-height"])for(var r,i=0;r=e[i];i++)if(r.contentWindow===a.source){var d=a.data["datawrapper-height"][t]+"px";r.style.height=d}}}))}();
</script>

Excluding a recent rise in overall enrollment after the COVID-19 pandemic — in the past 12 years, enrollment for Black students has steadily decreased as Hispanic enrollment has stayed relatively steady. This has resulted in a student body with a heavier Hispanic/Latinx presence.

In the 1980s, Black enrollment at the City Colleges of Chicago was substantially higher than any other group of race/ethnicity, according to data from the Illinois Community College Board. By the late 1990s/early 2000s, enrollment numbers between Black and Hispanic students started to even out. 2012 was the crossing point. After 2012, Black enrollment declined lower than than the Hispanic enrollment.

The decline in Black enrollment reflected the decline in total enrollment at CCC, which could be caused by several factors — changes in city demographic makeup, the development of more colleges/higher education options, etc. However, as the data shows later, the percent change in Black enrollment has higher than the total percent change, meaning Black enrollment declined at a steeper rate.

Combining this data with interviews and historical context can help us understand how the accessibility/use of these colleges for students of color in the area has developed.

<iframe title="Fall enrollment across City Colleges of Chicago for Black and Hispanic/Latinx students" aria-label="Grouped Columns" id="datawrapper-chart-xxniH" src="https://datawrapper.dwcdn.net/xxniH/5/" scrolling="no" frameborder="0" style="width: 0; min-width: 100% !important; border: none;" height="517" data-external="1"></iframe><script type="text/javascript">!function(){"use strict";window.addEventListener("message",(function(a){if(void 0!==a.data["datawrapper-height"]){var e=document.querySelectorAll("iframe");for(var t in a.data["datawrapper-height"])for(var r,i=0;r=e[i];i++)if(r.contentWindow===a.source){var d=a.data["datawrapper-height"][t]+"px";r.style.height=d}}}))}();
</script>

<iframe title="Changes in Black student enrollment compared to college totals." aria-label="Grouped Columns" id="datawrapper-chart-Zp5aq" src="https://datawrapper.dwcdn.net/Zp5aq/5/" scrolling="no" frameborder="0" style="width: 0; min-width: 100% !important; border: none;" height="500" data-external="1"></iframe><script type="text/javascript">!function(){"use strict";window.addEventListener("message",(function(a){if(void 0!==a.data["datawrapper-height"]){var e=document.querySelectorAll("iframe");for(var t in a.data["datawrapper-height"])for(var r,i=0;r=e[i];i++)if(r.contentWindow===a.source){var d=a.data["datawrapper-height"][t]+"px";r.style.height=d}}}))}();
</script>

Looking into this shift in community college enrollment is important because of the disproportionately high percentages of minority students who attend them. According to a 2022 study by the American Association of Community Colleges, nearly half of all U.S. students of color who were enrolled in postsecondary education attended a community college. 

Especially as DEI initiatives and grants for students of color are being threatened, it is timely to take a closer look at the institutions that serve as a more affordable option and if they are truly serving the students who need them. A study by UIUC's Office of Community College Research and Leadership found that although community colleges are often the most accessible postsecondary path for minority students, it is also one of the most chronically under-resourced sector of higher education in the U.S. It would be interesting to explore this lack of resources further by looking at funding, teacher qualifications, student to teacher ratios and retention/graduation rates.

In each of CCC's seven colleges, students of color make up over 60% of the student enrollment. By compiling the data from the past 20 years, we can see if the percentages of Black and Latinx students have always been high or if they have increased over the years. The Investigative Project on Race and Equity was originally interested in looking at acceptances and rejections across zip codes. However, all seven colleges are open enrollment, meaning anyone who applies is awarded admission.

We can also compare CCC's growth with the population growth of the city to see if the two align.

<iframe title="Chicago population over time" aria-label="Table" id="datawrapper-chart-lPBnm" src="https://datawrapper.dwcdn.net/lPBnm/1/" scrolling="no" frameborder="0" style="width: 0; min-width: 100% !important; border: none;" height="243" data-external="1"></iframe><script type="text/javascript">!function(){"use strict";window.addEventListener("message",(function(a){if(void 0!==a.data["datawrapper-height"]){var e=document.querySelectorAll("iframe");for(var t in a.data["datawrapper-height"])for(var r,i=0;r=e[i];i++)if(r.contentWindow===a.source){var d=a.data["datawrapper-height"][t]+"px";r.style.height=d}}}))}();
</script>

From looking at the city's population changes, we can see that as a whole, the Black population is decreasing and the Hispanic population is increasing. This could be due to many factors. Many news organizations have reported on the trend of out-migration of Black families from Chicago, leading to demographic shifts. 

Besides these demographic changes, the decreased enrollment in CCC could be due to the rise of other institutions that offer similar services or CCC's inability to serve existing needs.

Before finding data in Illinois Community College Board's data books, I spent some time looking at the statistical digests included in CCC's annual budgets. This data was very messy and many of the numbers did not add up, so I wouldn't want to cite numbers from those reports. The data was also only limited to students pursuing a degree program. However, my overall findings from the digests matched my conclusions from the college board data, which helped to verify the information. One of the takeaways I gathered through the statistical digests was that from FY 2017 to FY 2023, the enrollment numbers for Black students decreased slightly, while the numbers for Latinx students increased. In the 2022-23 academic year, which was the most recent year available, 7,722 students enrolled in a degree program. Of those, 27.5% were Black and 48.1% were Latinx. In 2017, the degree enrollment was similar at 7,224 students. However, 33.4% were Black and 37.7% were Latinx. 

Because CCC has a long history, for a more personal angle, we could try to find older POC alumni who attended the colleges years ago to compare to the experiences of students of color now. Students who have attended CCC would care about the article, especially if they have feelings on the cultural/racial representation within the colleges. We could do this by searching for relevant Facebook groups or directories.

<iframe title=" Black student enrollment drops in City Colleges of Chicago as  Latinx student enrollment stays steady. 
" aria-label="Interactive line chart" id="datawrapper-chart-iHTUw" src="https://datawrapper.dwcdn.net/iHTUw/3/" scrolling="no" frameborder="0" style="width: 0; min-width: 100% !important; border: none;" height="515" data-external="1"></iframe><script type="text/javascript">!function(){"use strict";window.addEventListener("message",(function(a){if(void 0!==a.data["datawrapper-height"]){var e=document.querySelectorAll("iframe");for(var t in a.data["datawrapper-height"])for(var r,i=0;r=e[i];i++)if(r.contentWindow===a.source){var d=a.data["datawrapper-height"][t]+"px";r.style.height=d}}}))}();
</script>

## Records/data 

The Illinois Community College Board has data on fall enrollment available dating back to 1983. I compiled enrollment data by ethnicity from 2000 to 2024 to better understand how enrollment has changed.

This data was much more comprehensive than the data provided in City Colleges of Chicago's statistical digests. It included students who were not seeking degrees, which the statistical digests did not. However, my findings on the decline in Black enrollment and steadiness of Latinx enrollment were the same for both.

College Scorecard also provides snapshots on each community college, including graduation rate, annual costs and median household income, which is good context. I used this in my first graphic to provide general overviews of each college.

## People I can interview 

- Lorenzo Baber is the director of UIUC's Office of Community College Research and Leadership. He or someone on the research team could provide insight on persisting equity issues within community colleges. He and Heather McCambly were also authors on a paper titled, 'Community Colleges as Racialized Organizations: Outlining Opportunities for Equity,' which would be very relevant.

- In April, Chicago Public Schools, CCC and the University of Illinois Chicago launched “Chicago Roadmap 2.0,” which will try to create a pipeline for students seeking higher education. The website does not include the directors of the project, but I could reach out to the general email to have someone on the project speak about the local need for students in the city.

- A researcher/historian at the Abakanowicz Research Center within the Chicago History Museum could provide context on the demographic development of the city and the rise in Latinx population.

- Young and old alumni of the colleges who are people of color could show the range of student experiences.

## Other visuals

I would like to insert a chloropleth map of the Hispanic population of Chicago in 2020, 2010 and 2000. I am just not sure how to get the census data to generate on the map. I could also do a chloropleth map showing changes in household income over time.

If it would be possible to find real people who had gone through a City of Columbia College years ago, it would be cool to have photos of them in college and them now. I could look through yearbooks to find sources.

## Realistic deadline 

I would likely need three weeks to finish this project in order to analyze my code and have enough time to schedule/conduct interviews.

## Other threads to follow

To flesh out the story, I think it would also be interesting to look into Chicago Public Schools enrollment, to see if there are similar trends. An [article by Illinois Policy](https://www.illinoispolicy.org/chicago-public-schools-sees-uptick-in-enrollment-but-still-lower-than-10-years-ago/) reported that Black students have recorded the largest decrease in enrollment since 2014. The article was published in 2024 and said there were 44,680 fewer Black students enrolled in CPS in 2024 than 10 years prior, marking a decline of 29% during 10 years.

Another thread that would be interesting to follow is if community colleges have the resources to support growth in Hispanic enrollment. An [article by WGNTV](https://wgntv.com/migrants-in-chicago/chicago-public-school-teachers-allegedly-told-to-pass-migrant-students-regardless-of-grades-report/) found that CPS teachers were allegedly told to pass migrant students regardless of grades. Because of the lack of resources at community colleges, it would be interesting to look into the ESL opportunities and the breadth of English language learning materials offered.

### Load software libraries
```{r}
#install.packages("tidyverse")
#install.packages("janitor")
#install.packages("readxl")
#install.packages("rvest")
```

```{r, echo = FALSE}
library(tidyverse)
library(janitor)
library(readxl)
library(rvest)

```


### Loading data:

```{r}
enroll <- read.csv("../data/Fall Enrollment CCC Since 2000 - Sheet1 (1).csv") |>
  janitor::clean_names()
```

### Data analysis:

Let's first take a look at the breakdown of students by college:
```{r, echo = FALSE}
library(tidyverse)
library(janitor)
library(knitr)

# Fall 2024 breakdown of enrollment by college and ethnicity (wide format)
fall_2024_breakdown <- enroll |>
  filter(fall_year == "2024") |>
  select(
    college,
    asian,
    native_american,
    african_american_non_latino,
    latino,
    white_non_latino,
    u_s_nonresident,
    pacific_islander,
    middle_eastern_or_north_african,
    two_or_more_races,
    all_other_or_no_indication,
    total
  ) |>
  rename(
    african_american = african_american_non_latino,
    mid_east_african = middle_eastern_or_north_african
  ) |>
  mutate(across(
    .cols = -college,
    .fns = ~ as.numeric(str_replace_all(as.character(.), ",", ""))
  )) |>
  clean_names()

# Show the table
kable(fall_2024_breakdown, caption = "Fall 2024 Enrollment by College and Ethnicity")

```
This table shows that for six of the colleges, African American and Latino students have the highest enrollment. The only outlier is Wilbur Wright.

Using this information, we can look at the percentages of Black and Latino students. It would also be interesting to know if this breakdown has always been the case. 
```{r}
enroll <- enroll |>
  mutate(
    asian = as.numeric(str_replace_all(asian, ",", "")),
    native_american = as.numeric(str_replace_all(native_american, ",", "")),
    african_american_non_latino = as.numeric(str_replace_all(african_american_non_latino, ",", "")),
    latino = as.numeric(str_replace_all(latino, ",", "")),
    white_non_latino = as.numeric(str_replace_all(white_non_latino, ",", "")),
    total = as.numeric(str_replace_all(total, ",", "")),
    asian_pct = (asian / total) * 100,
    native_american_pct = (native_american / total) * 100,
    african_american_non_latino_pct = (african_american_non_latino / total) * 100,
    latino_pct = (latino / total) * 100,
    white_non_latino_pct = (white_non_latino / total) * 100
  )

```
Let's look at how the percentage of Black students in each college has changed since 2014:
```{r}
library(tidyverse)
library(janitor)
library(knitr)

makeup_black <- enroll |>
  select(fall_year, college, african_american_non_latino_pct) |>
  filter(fall_year == "2024" | fall_year == "2014") |>
  pivot_wider(names_from = fall_year, values_from = african_american_non_latino_pct, id_cols = college) |>
  clean_names() |>
  mutate(
    x2024 = round(as.numeric(str_replace_all(x2024, ",", "")), 2),
    x2014 = round(as.numeric(str_replace_all(x2014, ",", "")), 2)
  )

 #Show the table
kable(makeup_black, caption = "Breakdown of Black students in 2014 and 2024 by percentage")
```

Now let's look at the change in the percentage of Latinx student since 2014 across each college:
```{r}
library(tidyverse)
makeup_latinx_2014 <- enroll |>
  select(fall_year, college, latino_pct) |>
  filter(fall_year == "2024" | fall_year == "2014") |>
  pivot_wider(names_from = fall_year, values_from = latino_pct, id_cols = college) |>
  janitor::clean_names() |>
mutate(
    x2024 = round(as.numeric(str_replace_all(x2024, ",", "")), 2),
    x2014 = round(as.numeric(str_replace_all(x2014, ",", "")), 2)
  )

# Show the table
kable(makeup_latinx_2014, caption = "Breakdown of Latinx students in 2014 and 2024 by percentage")
```

The results show that excluding the percentage of Latinx students who make up the student body have increased substantially across all colleges. Latinx students account for more than half of the student body in three out of the seven schools.

#### Q: What was the percent change in fall enrollment between 2024 and 2014 for CCC as a whole?

```{r}
library(tidyverse)
library(janitor)
library(knitr)

# Black student enrollment change
black_enrollment_10years <- enroll |>
  select(fall_year, college, african_american_non_latino) |>
  filter(fall_year == "2024" | fall_year == "2014") |>
  pivot_wider(names_from = fall_year, values_from = african_american_non_latino, id_cols = college) |>
  clean_names() |>
  mutate(
    x2024 = as.numeric(str_replace_all(x2024, ",", "")),
    x2014 = as.numeric(str_replace_all(x2014, ",", "")),
    pct_change = ((x2024 - x2014) / x2014)*100
  )|>
  arrange(desc(pct_change))

# Show the black enrollment table
kable(black_enrollment_10years, caption = "Change in Black Enrollment by College")

# Overall enrollment change
enroll1 <- enroll |>
  select(fall_year, college, total) |>
  filter(fall_year == "2024" | fall_year == "2014") |>
  pivot_wider(names_from = fall_year, values_from = total, id_cols = college) |>
  clean_names() |>
  mutate(
    x2024 = as.numeric(str_replace_all(x2024, ",", "")),
    x2014 = as.numeric(str_replace_all(x2014, ",", "")),
    pct_change = ((x2024 - x2014) / x2014)*100
  )|>
  arrange(desc(pct_change))

# Show the total enrollment table
kable(enroll1, caption = "Change in Total Enrollment by College")
```

```
Comparing the percent change between Black students and total students, we can see that Black student enrollment declined more steeply than total enrollment.

#### Q: What is the percent change in fall enrollment between 2024 and 2004 for Black students?


```{r, echo=FALSE}
library(tidyverse)
library(janitor)
library(knitr)

black_enrollment_20years <- enroll |>
  select(fall_year, college, african_american_non_latino) |>
  filter(fall_year == "2024" | fall_year == "2004") |>
  pivot_wider(names_from = fall_year, values_from = african_american_non_latino, id_cols = college) |>
  clean_names() |>
  mutate(
    x2024 = as.numeric(str_replace_all(x2024, ",", "")),
    x2004 = as.numeric(str_replace_all(x2004, ",", "")),
    pct_change = round(((x2024 - x2004) / x2004) * 100, 2)
  )|>
  arrange(desc(pct_change))

  
  # Show the total enrollment table
kable(black_enrollment_20years, caption = "Percent change in fall enrollment between 2024 and 2004 for Black students")

```

#### Q: What is the percent change in fall enrollment between 2024 and 2014 for Latinx students?
```{r, echo = FALSE}
library(tidyverse)
library(janitor)
library(knitr)

latinx_enrollment_10years <- enroll |>
  select(fall_year, college, latino) |>
  filter(fall_year == "2024" | fall_year == "2014") |>
  pivot_wider(names_from = fall_year, values_from = latino, id_cols = college) |>
  clean_names() |>
  mutate(
    x2024 = as.numeric(str_replace_all(x2024, ",", "")),
    x2014 = as.numeric(str_replace_all(x2014, ",", "")),
    pct_change = round(((x2024 - x2014) / x2014) * 100, 2)
  )|>
  arrange(desc(pct_change))
  
  # Show the total enrollment table
kable(latinx_enrollment_10years, caption = "Percent change in fall enrollment between 2024 and 2014 for Latinx students")

```
#### Q: What is the percent change in fall enrollment between 2024 and 2004 for Latinx students?
```{r, echo = FALSE}

library(tidyverse)
library(janitor)
library(knitr)

latinx_enrollment_20years <- enroll |>
  select(fall_year, college, latino) |>
  filter(fall_year == "2024" | fall_year == "2004") |>
  pivot_wider(names_from = fall_year, values_from = latino, id_cols = college) |>
  clean_names() |>
  mutate(
    x2024 = as.numeric(str_replace_all(x2024, ",", "")),
    x2004 = as.numeric(str_replace_all(x2004, ",", "")),
    pct_change = round(((x2024 - x2004) / x2004) * 100, 2)
  )|>
  arrange(desc(pct_change))
  
    # Show the total enrollment table
kable(latinx_enrollment_20years, caption = "Percent change in fall enrollment between 2024 and 2004 for Latinx students")

```




